Unknown

### 2025-06-22 ms/ms.qmd
working on ms

<!-- Paragraph 2 -->
Despite extensive theoretical and empirical progress understanding stomata function and anatomy from molecular to ecosystem levels, the adaptive significance of amphistomatous leaves remains an important unsolved problem in leaf structure-function relationships [@grubb_leaf_1977; @parkhurst_adaptive_1978; @mott_adaptive_1982;@gibson_structure_1996; @oguchi_leaf_2018; @drake_two_2019; @grubb_leaf_2020]. Amphistomatous leaves develop abaxial and adaxial stomata that can be independently regulated [@mott_stomatal_1984; @reich_effects_1985; @mott_asymmetric_1993] (reich, mott) to allow gas exchange through each surface (wall, marquez). In theory, gas exchange through stomata on each surface increases CO$_2$ supply to chloroplasts, all else being equal, enhancing photosynthesis (parkhurst, gutschick, fig. 1). Amphistomatous leaves also lose more water through evaporation because of a second boundary layer conductance (Smith), but the additional carbon gain should be enough to offset this cost in most realistic scenarios (Muir 2019).

<!-- Paragraph 3 -->
The paradoxical fact is that, despite the potential photosynthetic benefit, most leaves are not amphistomatous. Many vertically oriented and/or isobilateral leaves are amphistomatous (arcto paper, drake). But among dorsiventral leaves, herbaceous plants in open, high light habitats tend to have have amphistomatous leaves (salisbury, bucher, jordan, etc.). Most other leaves, except those from aquatic habitats, are hypostomatous, producing stomata only on the lower, abaxial surface. Even resupinate leaves develop stomata on the lower, albeit adaxial surface (Lesheyde?), suggesting that leaf orientation (lower vs. upper) rather than leaf polarity (abaxial vs. adaxial) is causal. The covariation between stomatal density ratio and light habitat is both qualitative and quantitative. A higher proportion of sun leaves are amphistomatous (salisbury) and the proportion of stomata on the upper, adaxial surface increases with light (muir 2018, bucher?). Resolving why high light intensity favors amphistomatous dorsivental leaves is an important first step toward understanding variation in stomatal density ratio and leaf structure-function relationships more generally.

@@

>>>>

<!-- Paragraph 2 -->
Despite extensive theoretical and empirical progress understanding stomata function and anatomy from molecular to ecosystem levels, the adaptive significance of amphistomatous leaves remains an important unsolved problem in leaf structure-function relationships [@grubb_leaf_1977; @parkhurst_adaptive_1978; @mott_adaptive_1982;@gibson_structure_1996; @oguchi_leaf_2018; @drake_two_2019; @grubb_leaf_2020]. Amphistomatous leaves develop abaxial and adaxial stomata that can be independently regulated [@pospisilova_stomatal_1980; @mott_stomatal_1984; @reich_effects_1985; @mott_asymmetric_1993; @wall_stomata_2022] to allow gas exchange through each surface. In theory, gas exchange through stomata on each surface increases CO$_2$ supply to chloroplasts, all else being equal, enhancing photosynthesis [@parkhurst_adaptive_1978; @gutschick_photosynthesis_1984]. Amphistomatous leaves also lose more water through evaporation because of a second boundary layer conductance (Smith), but the additional carbon gain should be enough to offset this cost in most realistic scenarios (Muir 2019).

<!-- Paragraph 3 -->
<!-- Need to mention here or in previous grafs, estimates of r_ias Mott & Oleary 1984; Marquez et al., etc. -->
<!-- "Farquhar and Raschke (2). However, using a technique similar to ours, Sharkey et al. (11) estimate r, in Xan:hium to be closer to 1.0 m2 s mol-', and state that they believe the estimate of 3.2 m2 s mol-' to be high" -->

The paradoxical fact is that, despite the potential photosynthetic benefit, most leaves are not amphistomatous. Many vertically oriented and/or isobilateral leaves are amphistomatous (arcto paper, drake). But among dorsiventral leaves, herbaceous plants in open, high light habitats tend to have have amphistomatous leaves (salisbury, bucher, jordan, etc.). Most other leaves, except those from aquatic habitats, are hypostomatous, producing stomata only on the lower, abaxial surface. Even resupinate leaves develop stomata on the lower, albeit adaxial surface (Lesheyde?), suggesting that leaf orientation (lower vs. upper) rather than leaf polarity (abaxial vs. adaxial) is causal. The covariation between stomatal density ratio and light habitat is both qualitative and quantitative. A higher proportion of sun leaves are amphistomatous (salisbury) and the proportion of stomata on the upper, adaxial surface increases with light (muir 2018, bucher?). Resolving why high light intensity favors amphistomatous dorsivental leaves is an important first step toward understanding variation in stomatal density ratio and leaf structure-function relationships more generally.

@@

from Mott and Oleary: "However, Jones and Slatyer (5) reported a higher mesophyll resistance for CO, entering through the upper stomata than for the lower, and the data of Vaclavik (12) appear to support this conclusion"
